---
title: "Proper Data Analysis"
output: 
  html_notebook:
    toc: True
    toc_float: True
---
## Introduction
Through the help of [Sharp Sight Labs](http://sharpsightlabs.com/), I will be able to propely analyze datasets by extracting, examining, cleaning, and merging data.

### Required Packages
```{r eval=FALSE}
require(tidyverse)
```

## Extracting
So the first thing to do of course is extracting data from a source. This won't always be as easy as pulling data from a given online source. YOu might be required to scrape data on your own in the future. But in this case, we will use data that was already scraped into .txt files using the read.csv function.
```{r}
df.car_torque <- read.csv(url("http://www.sharpsightlabs.com/wp-content/uploads/2014/11/auto-snout_torque_DATA.txt"))
df.car_0_60_times  <- read.csv(url("http://www.sharpsightlabs.com/wp-content/uploads/2014/11/auto-snout_0-60-times_DATA.txt"))
df.car_engine_size <- read.csv(url("http://www.sharpsightlabs.com/wp-content/uploads/2014/11/auto-snout_engine-size_DATA.txt"))
df.car_horsepower  <- read.csv(url("http://www.sharpsightlabs.com/wp-content/uploads/2014/11/auto-snout_horsepower_DATA.txt"))
df.car_top_speed   <- read.csv(url("http://www.sharpsightlabs.com/wp-content/uploads/2014/11/auto-snout_top-speed_DATA.txt"))
df.car_power_to_weight <- read.csv(url("http://www.sharpsightlabs.com/wp-content/uploads/2014/11/auto-snout_power-to-weight_DATA.txt"))
```
Each url was a data frames assigned to a name according to the appropriate characteristics of variables within them. We can take a look at any one. Next, using the head() function from dplyr, we can inspect the first couple of rows of a dataset. For example: let's take a look at the dataset that display specific vehicles and the time it takes the vehicles to make it from 0-60 seconds.

```{r}
head(df.car_horsepower)
```
## Examining Data
The next step would be to examine the data for and duplicate information. Thr reason we do this is so that we can prevent and errors when we join (combine) datasets in the future. We will once again be using dplyr functions to isolate these duplications. The functions group_by and summarise will divide the dataset into individual groups. In this case we use these the two functions to examine the full name of a car and analyzing the duplication of the data using filter().
```{r}
df.car_torque      %>% group_by(car_full_nm) %>% summarise(count=n()) %>% filter(count!=1)
df.car_0_60_times  %>% group_by(car_full_nm) %>% summarise(count=n()) %>% filter(count!=1)
df.car_engine_size %>% group_by(car_full_nm) %>% summarise(count=n()) %>% filter(count!=1)
df.car_horsepower  %>% group_by(car_full_nm) %>% summarise(count=n()) %>% filter(count!=1)
df.car_top_speed   %>% group_by(car_full_nm) %>% summarise(count=n()) %>% filter(count!=1)
df.car_power_to_weight %>% group_by(car_full_nm) %>% summarise(count=n()) %>% filter(count!=1)
```
Now as you see, we have filter the dataset to see which cars were duplicated. This process is known as __aggregating__. We'll now use the distinct() function to eliminate duplications in each dataset.
```{r}
df.car_0_60_times  <- distinct(df.car_0_60_times ,car_full_nm,.keep_all = TRUE)
df.car_engine_size <- distinct(df.car_engine_size ,car_full_nm,.keep_all = TRUE)
df.car_horsepower  <- distinct(df.car_horsepower ,car_full_nm ,.keep_all = TRUE)
df.car_top_speed   <- distinct(df.car_top_speed ,car_full_nm,.keep_all = TRUE)
df.car_torque      <- distinct(df.car_torque ,car_full_nm,.keep_all = TRUE)
df.car_power_to_weight <- distinct(df.car_power_to_weight, car_full_nm,.keep_all = TRUE)
```

## Joining
Now that we have examined each dataframe for duplications and eliminated them, we can now join the data together without any faulty joins. I proceed to join the horsepower dataframe and the torque dataframe to the car_full_nm variable. After this we can simply continue to modify our (let's call it) car_spec_data dataframe so that all of our data becomes one dataframe. One thing to evaluate as we continue to modify our data is the number of records. This should be the same as we continue each step. If the number increases, that means that there are still duplications and we'll have to go back and fix our functions.
```{r eval=FALSE}
df.car_spec_data <- left_join(df.car_horsepower, df.car_torque, by="car_full_nm")     
df.car_spec_data <- left_join(df.car_spec_data, df.car_0_60_times, by="car_full_nm")  
df.car_spec_data <- left_join(df.car_spec_data, df.car_engine_size, by="car_full_nm") 
df.car_spec_data <- left_join(df.car_spec_data, df.car_top_speed, by="car_full_nm")   
df.car_spec_data <- left_join(df.car_spec_data, df.car_power_to_weight, by="car_full_nm")
```

Now we can test the dataframe to see if there are any duplicates.
```{r}
df.car_spec_data      %>% group_by(car_full_nm) %>% summarise(count=n()) %>% filter(count!=1)
```
Luckily, there are none. Let's take a look at the dataframe.
```{r}
str(df.car_spec_data)
head(df.car_spec_data)
```

## Mutating
After we inspected the data, we will now add variables to the whole dataframe. In this section we'll use the dplyr functions mutate() and sub(). Mutate simply adds the variable(s) you want to the data frame, add a new column with differnet values. Sub() parses character vector and replaces part of a string with something else. For this data frame, we want to add the variable of time. Let's add the model year of each car and the decade they came out.
```{r}
# NEW VAR: year
df.car_spec_data <- mutate(df.car_spec_data, year=sub(".*\\[([0-9]{4})\\]","\\1",car_full_nm))

str(df.car_spec_data$year)
```
```{r}
# NEW VAR: decade

df.car_spec_data <- mutate(df.car_spec_data, 
                           decade = as.factor(
                                    ifelse(substring(df.car_spec_data$year,1,3)=='193','1930s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='194','1940s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='195','1950s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='196','1960s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='197','1970s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='198','1980s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='199','1990s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='200','2000s',
                                    ifelse(substring(df.car_spec_data$year,1,3)=='201','2010s',"ERROR"
                                    )))))))))
                                    )
                            )
head(df.car_spec_data)
str(df.car_spec_data)
```
What I have done in the last chunk is classified each car within year containing 193 as a 1930s' car and so on all the way up to the 201's or 2010 decade. I will also add other variables like the brand of the car, the weight of each car, and the torque per ton. Each of these variables use the values of other variables to calculate the values of the new variables. Take a look for yourself.

```{r}
# NEW VAR: brand_nm 

df.car_spec_data <- mutate(df.car_spec_data, brand_nm = gsub(" .*$","", df.car_spec_data$car_full_nm))
```
```{r}
# NEW VAR: car_weight_tons 

df.car_spec_data <- mutate(df.car_spec_data, car_weight_tons = horsepower_bhp/ horsepower_per_ton_bhp)
```
```{r}
# NEW VAR: torque_per_ton 

df.car_spec_data <- mutate(df.car_spec_data, torque_per_ton = torque_lb_ft / car_weight_tons)
```
## Inspection
After adding all of these new variables the wisest thing to do at this moment would be to inspect the data frame again to make sure everything variable added came out smoothly.
```{r}
head(df.car_spec_data)
```
WE know that our new variables have be added to the data frame but we can still evaluate the individual variable on their own and find out how much of what is where.
```{r}
#Decade
df.car_spec_data %>%  
  group_by(decade) %>%
  summarize(count=n())
```
```{r}
#Brand
df.car_spec_data %>%
  group_by(brand_nm) %>%
  summarise(brand_count = length(brand_nm)) %>%
  arrange(desc(brand_count))
```
## The Exploration
Now that we have our offfcial data frame, we can now graph what we want to see within the dataset using ggplot2 and ofcourse, dplyr. These are the tools that will be use to highlight any trends within the data. First thing's first. We will make a theme for our graphs. These are just a bunch of pre-set setting and technical aesthetics. We set them up for future use:
```{r}
# Basic Theme
theme.car_chart <- 
  theme(legend.position = "none") +
  theme(plot.title = element_text(size=26, family="Trebuchet MS", face="bold", hjust=0, color="#666666")) +
  theme(axis.title = element_text(size=18, family="Trebuchet MS", face="bold", color="#666666")) +
  theme(axis.title.y = element_text(angle=0)) 
# SCATTERPLOT THEME
theme.car_chart_SCATTER <- theme.car_chart +
                            theme(axis.title.x = element_text(hjust=0, vjust=-.5))

# HISTOGRAM THEME
theme.car_chart_HIST <- theme.car_chart +
                          theme(axis.title.x = element_text(hjust=0, vjust=-.5))

# SMALL MULTIPLE THEME
theme.car_chart_SMALLM <- theme.car_chart +
                            theme(panel.grid.minor = element_blank()) +
                            theme(strip.text.x = element_text(size=16, family="Trebuchet MS", face="bold", color="#666666"))
```
### Speed
Let's analyze Horsepower vs. Top Speed
```{r}
ggplot(data=df.car_spec_data, aes(x=horsepower_bhp, y=top_speed_mph)) +
  geom_point(alpha=.4, size=4, color="#880011") +
  ggtitle("Horsepower vs. Top Speed") +
  labs(x="Horsepower, bhp", y="Top Speed,\n mph") +
  theme.car_chart_SCATTER
```
As you can see, horsepower and top speed are definetly related. You can also see a bunch of car that have a top speed of around 150 mph. According to the [Sharp Sight Labs](http://sharpsightlabs.com/blog/data-analysis-example-r-supercars-part2/) websites, this can be due to speed limiter systems. To futher evaluate this theory, we can create a histogram that shows use the amount of cars the have top speeds of around 150 mph.
```{r}
ggplot(data=df.car_spec_data, aes(x=top_speed_mph)) +
  geom_histogram(fill="#880011") +  
  ggtitle("Histogram of Top Speed") +
  labs(x="Top Speed, mph", y="Count\nof Records") +
  theme.car_chart_HIST
```
Even though this graph is a little more specific to our cause, it's still not pinpointing that major spike containing vehicles with top speeds from 150-160 mph. So let's subset out data for speeds within that range and graph it into a bar graph.

```{r}
#SPEED LIMITED CARS
df.car_spec_data %>%
  filter(top_speed_mph >149 & top_speed_mph <159) %>%
  ggplot(aes(x= as.factor(top_speed_mph))) +
    geom_bar(fill="#880011") +
    labs(x="Top Speed, mph") +
    theme.car_chart
```
We can continue speculating that the reason for so many cars having a top speed of 155 mph  is due to that speed limiting factor. Next, we can evaluate the top speeds for cars across decades using the facet technique.
```{r}
ggplot(data=df.car_spec_data, aes(x=top_speed_mph)) +
  geom_histogram(fill="#880011") +
  ggtitle("Histogram of Top Speed\nby decade") +
  labs(x="Top Speed, mph", y="Count\nof Records") +
  facet_wrap(~decade) +
  theme.car_chart_SMALLM
```
Let's take a look at the companies that limited the top speeds of these cars made after 1990 using the same dplyr functions we discussed earlier.

```{r}
df.car_spec_data %>%
  filter(top_speed_mph == 155 & year>=1990) %>%
  group_by(brand_nm) %>% 
  summarize(count_speed_controlled = n()) %>%
  arrange(desc(count_speed_controlled))
```
### Visualization Cont.
Let's take a look at horsepower vs. speed by decade and in different facets.
```{r}
ggplot(data=df.car_spec_data, aes(x=horsepower_bhp, y=top_speed_mph)) +
  geom_point(alpha=.6,color="#880011") +
  facet_wrap(~decade) +
  ggtitle("Horsepower vs Top Speed\nby decade") +
  labs(x="Horsepower, bhp", y="Top Speed\n mph") +
  theme.car_chart_SMALLM
```
As you can see in the graph there are some cars that have high horsepower and yet their top speed is not as high. I think this indirect relationship is due to the weight of the cars make in the 60's to 2000's. The materials used to make cars in these time were much heavier than the innovative, lighter metals used to design cars today. Thus, the top speeds have increased over time. To prove this theory and take a deeper look we can plot thr top speed vs. year.
```{r}
ggplot(data=df.car_spec_data, aes(x=year, y=df.car_spec_data$top_speed_mph)) +
  geom_point(alpha=.35, size=4.5, color="#880011", position = position_jitter()) +
  scale_x_discrete(breaks = c("1950","1960","1970","1980","1990","2000","2010")) +
  ggtitle("Car Top Speeds by Year") +
  labs(x="Year" ,y="Top Speed\nmph") +
  theme.car_chart_SCATTER
```
We can even check out the fastest car each year to see if there is an increase.
```{r}
df.car_spec_data %>%
  group_by(year) %>%
  summarize(max_speed = max(top_speed_mph, na.rm=TRUE)) %>%
  ggplot(aes(x=year,y=max_speed,group=1)) + 
    geom_point(size=5, alpha=.8, color="#880011") +
    stat_smooth(method="auto",size=1.5) +
    scale_x_discrete(breaks = c("1950","1960","1970","1980","1990","2000","2010")) +
    ggtitle("Speed of Year's Fastest Car by Year") +
    labs(x="Year",y="Top Speed\n(fastest car)") +
    theme.car_chart_SCATTER
```

### From 0 to 60
Now we ill compare the time is take for a car to get from 0 to 60 mph and horsepower.
```{r}
ggplot(data=df.car_spec_data, aes(x=horsepower_bhp,y=car_0_60_time_seconds)) +
  geom_point()
```
A lot of the points are in line and on top of one another. We can fix this problem by useing a position function. In this case the function would be jitter.
```{r}
ggplot(data=df.car_spec_data, aes(x=horsepower_bhp,y=car_0_60_time_seconds)) +
  geom_point(position="jitter")
```
Now we are able to see those points clearly. Now let's properly analyze the trend by ploting a line graph to it too using the stat_smooth() function.
```{r}
ggplot(data=df.car_spec_data, aes(x=horsepower_bhp,y=car_0_60_time_seconds)) +
  geom_point(size=4, alpha=.7,color="#880011",position="jitter") +
  stat_smooth(method="auto",size=1.5) +
  ggtitle("0 to 60 times by Horsepower") +
  labs(x="Horsepower, bhp",y="0-60 time\nseconds") +
  theme.car_chart_SCATTER
```
We'll do the same thing for 0-60 mph vs. horsepower per tonne
```{r}
ggplot(data=df.car_spec_data, aes(x=horsepower_per_ton_bhp,y=car_0_60_time_seconds)) +
  geom_point(size=4, alpha=.5,color="#880011",position="jitter") +
  stat_smooth(method="auto",size=1.5) +
  ggtitle("0 to 60 times\nbyHorsepower-per-Tonne") +
  labs(x="Horsepower-per-tonne",y="0-60 time\nseconds") +
  theme.car_chart_SCATTER
```
Once again the rate of acceleration increases as horsepower per tonne increases. Let's make a similar chart for torque per tonne.
```{r}
ggplot(data=df.car_spec_data, aes(x=df.car_spec_data$torque_per_ton,y=car_0_60_time_seconds)) +
  geom_point(size=4, alpha=.5,color="#880011",position="jitter") +
  stat_smooth(method="auto",size=1.5) +
  ggtitle("0 to 60 times\nby Torque-per-Tonne") +
  labs(x="Torque-per-tonne",y="0-60 time\nseconds") +
  theme.car_chart_SCATTER
```
### Top Ten
The same trend seems to appear in the graph above as the x value increases, the y value should decrease. Now let's make our final graph displaying the top ten fastest cars in descending order.
```{r}
df.car_spec_data %>%
  select(car_full_nm,top_speed_mph) %>%
  filter(min_rank(desc(top_speed_mph)) <= 10) %>%
  arrange(desc(top_speed_mph)) %>%
  ggplot(aes(x=reorder(car_full_nm,top_speed_mph), y=top_speed_mph)) +
    geom_bar(stat="identity",fill="#880011") +
    coord_flip() +
    ggtitle("Top 10 Fastest Cars (through 2012)") +
    labs(x="",y="") +
    theme.car_chart +
    theme(axis.text.y = element_text(size=rel(1.5))) +
    theme(plot.title = element_text(hjust=1))
```
## Conclusion
What we have just done is a proper data analysis. We pulled, cleaned, examined and joined data to visualize and evaluate trends of different variables. If your ever stuck with data analysis, refer back to this.




